---
title: "biomeR: An R Interface to Biome.jl"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{biomeR Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# biomeR

biomeR is an R frontend that calls the Julia package [Biome.jl](https://github.com/clechartre/Biome.jl) through JuliaCall.

## Typical workflow

1. Clone [Biome.jl](https://github.com/clechartre/Biome.jl) (until it is registered)
2. Initialize Julia + Biome.jl with `biome_setup()`
3. Load input rasters in R
4. **Use BIOME4 PFTs** with `make_biome4_pftlist()` and `set_pft_constraint()` **or** build custom PFTs with `pft()`
5. Run the model with `run_biome()`, producing a NetCDF output

---

## Installation

### Requirements

- R (≥ 4.1 recommended)
- Julia (≥ 1.10 recommended)
- A local clone of Biome.jl

### Step 1: Install biomeR

```{r, eval = FALSE}
install.packages("remotes")
remotes::install_github("clechartre/biomeR")
```

### Step 2: Install Julia & JuliaCall

```{r, eval = FALSE}
install.packages("JuliaCall")
library(JuliaCall)
julia_setup(installJulia = TRUE, force = TRUE)
```

### Step 3: Point biomeR to your Biome.jl checkout

```{r, eval = FALSE}
library(biomeR)

biome <- biome_setup(
  project_dir = "/path/to/Biome.jl",
  pkg_check = TRUE
)
```

---

## Load input data

biomeR expects climate inputs as rasters (typically 12 monthly layers).
We use [CHELSA](https://www.chelsa-climate.org/datasets) climatologies here,
but any source on a consistent grid works.

```{r, eval = FALSE}
library(raster)

# Load climate rasters (12 months each)
temp <- brick("/path/to/temp_1981-2010.nc", varname = "temp")
prec <- brick("/path/to/prec_1981-2010.nc", varname = "prec")
clt  <- brick("/path/to/sun_1981-2010.nc",  varname = "sun") 

# Load soil rasters
whc  <- brick("/path/to/soils_55km.nc", varname = "whc")
ksat <- brick("/path/to/soils_55km.nc", varname = "Ksat")


nlayers(temp)  # should be 12

rasters <- list(temp = temp, prec = prec, clt = clt, whc = whc, ksat = ksat)
```

---

# Workflow 1: BIOME4 with built-in PFTs

The simplest approach: run BIOME4 with its original parametrisation.

## Create the BIOME4 PFT list

```{r, eval = FALSE}
pftlist <- make_biome4_pftlist(biome)
```

`make_biome4_pftlist()` returns an integer handle referencing the Julia-side
`PFTClassification`. You can inspect the 13 PFT names at any time:

```{r, eval = FALSE}
get_pft_names(pftlist)
#  [1] "TropicalEvergreen"             "TropicalDroughtDeciduous"
#  [3] "TemperateBroadleavedEvergreen" "TemperateDeciduous"
#  [5] "CoolConifer"                   "BorealEvergreen"
#  [7] "BorealDeciduous"               "C3C4TemperateGrass"
#  [9] "C4TropicalGrass"               "C3C4WoodyDesert"
# [11] "TundraShrubs"                  "ColdHerbaceous"
# [13] "LichenForb"
```

## Edit PFT constraints (optional)

Use `set_pft_constraint()` to adjust any climatic constraint before running:

```{r, eval = FALSE}
set_pft_constraint(pftlist, "CoolConifer",    "gdd5", c(500.0, 1200.0))
set_pft_constraint(pftlist, "BorealEvergreen", "tcm", c(-32.5, -2.0))
```

## Run the model

```{r, eval = FALSE}
outfile <-  "biome_output_biome4.nc"

# Define a bounding box — returns an integer handle
ch_bounds <- bbox(5.9, 10.5, 45.8, 47.8)

# Pass to run_biome
run_biome(
  rasters = rasters,
  model   = "BIOME4Model",
  co2     = 373.8,
  pftlist = pftlist,
  bounds  = ch_bounds,
  outfile = outfile
)
```

---

# Workflow 2: Custom PFTs

Build PFTs entirely in R and send them to Julia.

## Create PFTs

```{r, eval = FALSE}
cool_conifer <- pft(
  type = "NeedleleafEvergreenPFT",
  name = "CoolConifer_custom",
  constraints = list(
    gdd5 = c(500, 1200),
    tcm  = c(-15, 2),
    twm  = c(5, 15)
  ),
  dominance_factor = 3L
)

boreal_dec <- pft(
  type = "NeedleleafDeciduousPFT",
  name = "BorealDeciduous_custom",
  constraints = list(
    gdd5 = c(300, 900),
    tcm  = c(-Inf, 5)
  ),
  dominance_factor = 3L
)
```

## Modify a PFT after creation

```{r, eval = FALSE}
cool_conifer_v2 <- modify_pft(cool_conifer, dominance_factor = 4L)
```

## Collect and run

```{r, eval = FALSE}
my_pfts <- pft_list(cool_conifer_v2, boreal_dec)

run_biome(
  rasters   = rasters,
  model     = "BaseModel",
  co2       = 380.0,
  pft_specs = my_pfts,
  outfile   = "biome_output_custom.nc"
)
```

---

# Complete end-to-end example

```{r, eval = FALSE}
library(biomeR)
library(raster)

# 1. Setup
biome <- biome_setup(project_dir = "/path/to/Biome.jl")

# 2. Load rasters
rasters <- list(
  temp = brick("temp.nc"),
  prec = brick("prec.nc"),
  clt  = brick("sun.nc"),
  whc  = brick("whc.nc"),
  ksat = brick("ksat.nc")
)

# 3. Build BIOME4 PFT list and edit as needed
pftlist <- make_biome4_pftlist(biome)
set_pft_constraint(pftlist, "CoolConifer",    "gdd5", c(500.0, 1200.0))
set_pft_constraint(pftlist, "BorealDeciduous", "tcm", c(-Inf,  5.0))

# 4. Run
run_biome(
  rasters = rasters,
  model   = "BIOME4Model",
  co2     = 373.8,
  pftlist = pftlist,
  outfile = "output.nc",
  biome   = biome
)
```

---

## Notes

- All rasters must be on the same grid.
- PFT `constraints` are named lists of length-2 numeric vectors.
- `make_biome4_pftlist()` returns an integer handle — pass it directly to
  `set_pft_constraint()` and `run_biome()`.
- Use `get_pft_names()` to inspect the PFTs in a list.
- Use `modify_pft()` to adjust custom PFT parameters after creation.
- Julia output is written to the NetCDF file specified by `outfile`.

For more details see the [Biome.jl documentation](https://github.com/clechartre/Biome.jl).
